\vspace{-\baselineskip} Fast and scalable Gaussian process modeling with applications to astronomical time series \vspace{-2\baselineskip}

### 2017-06-05 paper/ms.tex
some words

A source of variability that can be measured from time series measurements of
stars is rotation.
The inhomogeneous surface of the star (spots, plage, \etc) imprints itself as
quasiperiodic variations in photometric or spectroscopic observations
\citep{Dumusque:2014}.
It has been demonstrated that for light curves with nearly uniform sampling,
the empirical autocorrelation function provides a reliable estimate of the
rotation period of a star \citep{Mcquillan:2013, Mcquillan:2014, Aigrain:2015}
and that a GP model with a
quasiperiodic covariance function can be used to make probabilistic
measurements even with sparsely sampled data (R.~Angus, \etal\ submitted).
The covariance function used for this type of analysis has the form
\begin{eqnarray}\eqlabel{sine2}
k(\tau) = A\,\exp\left(-\frac{\tau^2}{2\,\ell^2} -
    \Gamma\,\sin^2\left(\frac{\pi\,\tau}{P} \right) \right)
\end{eqnarray}
where $P$ is the rotation period of the star.
GP modeling with the same kernel function has been proposed as a method of
measuring the mean periodicity in quasiperiodic photometric time series in
general \citep{Wang:2012}.
The key difference between \eq{sine2} and other quasiperiodic kernels is that
it is positive for all values of $\tau$.
We construct a simple \celeriteterm\ covariance function with similar
properties as follows
\begin{eqnarray}\eqlabel{rot-kernel}
k(\tau) = \frac{B}{2+C}\,e^{-\tau/L}\,\left[
    \cos\left(\frac{2\,\pi\,\tau}{P}\right) + (1 + C)
\right]
\end{eqnarray}
for $B>0$, $C>0$, and $L>0$.
The covariance function in \eq{rot-kernel} cannot exactly reproduce \eq{sine2}
but, since \eq{sine2} is only an effective model, \eq{rot-kernel} can be used
as a drop-in replacement for a significant gain in computational efficiency.

@@

We run \exampleiiinburn~steps of burn-in, followed by \exampleiiinsteps~steps
of MCMC using \emcee.
We estimate the integrated autocorrelation time of the chain for $\ln P$ and
estimate that we have \exampleiiineff\ independent samples.
These samples give a posterior constraint on the period of $P =
\rotationperiod$ and the marginalized posterior distribution for $P$ is shown
in the right panel of \Figure{rotation}.
This result is in good agreement with the literature value with smaller
uncertainties.
A detailed comparison of GP rotation period measurements and the traditional
methods is beyond the scope of this paper, but forthcoming work (R.~Angus,
\etal, submitted) demonstrates that GP inferences are, at a population level,
more reliable than other methods.
}

@@

\begin{figure}[p]
\begin{center}
\includegraphics[width=0.8\textwidth]{figures/rotation/rotation.pdf}
\caption{Inferred constraints on a quasiperiodic GP model using the covariance
    function in \eq{rot-kernel} and two quarters of \kepler\ data.
(top left) The \kepler\ data (black points) and the maximum likelihood model
    prediction (blue curve) for a 60~day subset of the data used.
    The solid blue line shows the predictive mean and the blue contours show
    the predictive standard deviation.
(bottom left) The residuals between the mean predictive model and the data
    shown in the top left figure.
(right) The posterior constraint on the rotation period of KIC~1430163 using
    the dataset and model from \Figure{rotation}.
    The period is the parameter $P$ in \eq{rot-kernel} and this figure shows
    the posterior distribution marginalized over all other nuisance parameters
    in \eq{rot-kernel}.
    This is consistent with the published rotation period made using the full
    \kepler\ baseline shown as a vertical gray line \citep{Mathur:2014}.
    \figurelabel{rotation}}
\end{center}
\end{figure}

@@

In \eq{gp-likelihood} the physical parameters of the exoplanet are called
$\bvec{\theta}$ and, in this example, the mean function $\mu_\bvec{\theta}(t)$
is a limb-darkened transit light curve \citep{Mandel:2002} and the parameters
$\bvec{\theta}$ are the orbital period $P_\mathrm{orb}$, the transit duration
$T$, the phase or epoch $t_0$, the impact parameter $B$, the radius of the
planet in units of the stellar radius $R_P/R_\star$, and two parameters
describing the limb-darkening profile of the star
\citep{Claret:2011,Kipping:2013}.
As in \sect{rotation}, we model the stellar variability using a GP model
with a kernel given by \eq{rot-kernel} and fit for the parameters of the
exoplanet $\bvec{\theta}$ and the stellar variability $\bvec{\alpha}$
simultaneously.
\response{The full set of parameters $\bvec{\alpha}$ and $\bvec{\theta}$ are
listed in Table~\ref{tab:example-5-params} along with their priors and the
true values for $\bvec{\theta}$.}

@@

\dfmtodo{Do this.}

>>>>

A source of variability that can be measured from time series measurements of
stars is rotation.
The inhomogeneous surface of the star (spots, plage, \etc) imprints itself as
quasiperiodic variations in photometric or spectroscopic observations
\citep{Dumusque:2014}.
It has been demonstrated that for light curves with nearly uniform sampling,
the empirical autocorrelation function provides a reliable estimate of the
rotation period of a star \citep{Mcquillan:2013, Mcquillan:2014, Aigrain:2015}
and that a GP model with a
quasiperiodic covariance function can be used to make probabilistic
measurements even with sparsely sampled data (R.~Angus, \etal\ submitted).
The covariance function used for this type of analysis has the form
\begin{eqnarray}\eqlabel{sine2}
k(\tau) = A\,\exp\left(-\frac{\tau^2}{2\,\ell^2} -
    \Gamma\,\sin^2\left(\frac{\pi\,\tau}{P_\mathrm{rot}} \right) \right)
\end{eqnarray}
where $P_\mathrm{rot}$ is the rotation period of the star.
GP modeling with the same kernel function has been proposed as a method of
measuring the mean periodicity in quasiperiodic photometric time series in
general \citep{Wang:2012}.
The key difference between \eq{sine2} and other quasiperiodic kernels is that
it is positive for all values of $\tau$.
We construct a simple \celeriteterm\ covariance function with similar
properties as follows
\begin{eqnarray}\eqlabel{rot-kernel}
k(\tau) = \frac{B}{2+C}\,e^{-\tau/L}\,\left[
    \cos\left(\frac{2\,\pi\,\tau}{P_\mathrm{rot}}\right) + (1 + C)
\right]
\end{eqnarray}
for $B>0$, $C>0$, and $L>0$.
The covariance function in \eq{rot-kernel} cannot exactly reproduce \eq{sine2}
but, since \eq{sine2} is only an effective model, \eq{rot-kernel} can be used
as a drop-in replacement for a significant gain in computational efficiency.

@@

We run \exampleiiinburn~steps of burn-in, followed by \exampleiiinsteps~steps
of MCMC using \emcee.
We estimate the integrated autocorrelation time of the chain for $\ln
P_\mathrm{rot}$ and estimate that we have \exampleiiineff\ independent
samples.
These samples give a posterior constraint on the period of $P_\mathrm{rot} =
\rotationperiod$ and the marginalized posterior distribution for $P$ is shown
in the right panel of \Figure{rotation}.
This result is in good agreement with the literature value with smaller
uncertainties.
A detailed comparison of GP rotation period measurements and the traditional
methods is beyond the scope of this paper, but forthcoming work (R.~Angus,
\etal, submitted) demonstrates that GP inferences are, at a population level,
more reliable than other methods.
}

@@

\begin{figure}[p]
\begin{center}
\includegraphics[width=0.8\textwidth]{figures/rotation/rotation.pdf}
\caption{Inferred constraints on a quasiperiodic GP model using the covariance
    function in \eq{rot-kernel} and two quarters of \kepler\ data.
(top left) The \kepler\ data (black points) and the maximum likelihood model
    prediction (blue curve) for a 60~day subset of the data used.
    The solid blue line shows the predictive mean and the blue contours show
    the predictive standard deviation.
(bottom left) The residuals between the mean predictive model and the data
    shown in the top left figure.
(right) The posterior constraint on the rotation period of KIC~1430163 using
    the dataset and model from \Figure{rotation}.
    The period is the parameter $P_\mathrm{rot}$ in \eq{rot-kernel} and this
    figure shows the posterior distribution marginalized over all other
    nuisance parameters in \eq{rot-kernel}.  This is consistent with the
    published rotation period made using the full
    \kepler\ baseline shown as a vertical gray line \citep{Mathur:2014}.
    \figurelabel{rotation}}
\end{center}
\end{figure}

@@

In \eq{gp-likelihood} the physical parameters of the exoplanet are called
$\bvec{\theta}$ and, in this example, the mean function $\mu_\bvec{\theta}(t)$
is a limb-darkened transit light curve \citep{Mandel:2002,
Foreman-Mackey:2016a} and the parameters $\bvec{\theta}$ are the orbital
period $P_\mathrm{orb}$, the transit duration $T$, the phase or epoch $t_0$,
the impact parameter $b$, the radius of the planet in units of the stellar
radius $R_P/R_\star$, the baseline relative flux of the light curve $f_0$, and
two parameters describing the limb-darkening profile of the star
\citep{Claret:2011, Kipping:2013}.
As in \sect{rotation}, we model the stellar variability using a GP model
with a kernel given by \eq{rot-kernel} and fit for the parameters of the
exoplanet $\bvec{\theta}$ and the stellar variability $\bvec{\alpha}$
simultaneously.
\response{The full set of parameters $\bvec{\alpha}$ and $\bvec{\theta}$ are
listed in Table~\ref{tab:example-5-params} along with their priors and the
true values for $\bvec{\theta}$.}

@@

\response{
In this section, we demonstrated five potential use cases for \celerite.
These examples span a range of data sizes and model complexities, and the
last three are all based on active areas of research with immediate real-world
applicability.
The sizes of these datasets are modest on the scale of some existing and
forthcoming survey datasets because we designed the experiments to be easily
reproducible but, even in these cases, the inferences made here would be
intractable without substantial computational investment.

Table~\ref{tab:example-stats} lists the specifications of each example.

In Example~3, for example, we inferred the rotation period of a star observed
by \kepler\ with a cost of about 4~CPU-minutes.
By contrast, the same inference using a general but optimized Cholesky
factorization routine would require nearly 400~CPU-hours.
The upcoming \tess\ Mission will observe \dfmtodo{XXX} stars.
}
